Toward Resolved Simulations of Burning Fronts in Thermonuclear X-ray Bursts

### 2018-10-03 astronum_2018/astronum-xrb.tex
a little smoothing of the ODE paragraph

\usepackage{amsmath}

\usepackage{bm}

@@

For the ODE integrations shown we use the variable-order,
variable-step integrator \vode\ \cite{vode} with its implementation of
the backward-differentiation formulas. The integration order
corresponds to the order of the interpolating polynomials used to
predict the solution in each integrator step. We note that for both
Strang splitting and SDC, \vode\ must start the integration at first
order and correspondingly small time steps since no higher-order
information about the solution is available. That is, our reaction
network right-hand sides provide the first time derivative of the
solution but no higher derivatives. Based on testing with a 13-isotope
approximate alpha-chain network, \vode\ takes many small timesteps
(e.g. $\approx 200$) before it can stably increase to a high order
representation of the solution, but after it does so, it can often
complete the integration in only a handful of steps (e.g.
$\approx 5-6$). Observing that during a \castro\ timestep, the first
Strang step is a direct continuation of the integration in the second
Strang step of the previous \castro\ timestep, we explored saving the
\vode\ state following the second Strang step and restarting \vode\
during the first Strang step. This approach allowed us to start the
first Strang step with a high order representation of the solution and
complete it in very few \vode\ timesteps (e.g. $<10$ steps), speeding
up the \castro\ timestep advance by 50\% for our detonation test
problem using the 13-isotope network. However, this came at the memory
cost of storing $7 N_{eq} + 90$ scalar variables on our grid, where
for our networks $N_{eq} = N_{spec} + 2$ (in addition to the number of
species $N_{spec}$ we also integrate temperature and specific
energy). For a large-scale science problem with a large network, this
could quickly become prohibitive, so we are exploring ways to mitigate
either the memory requirements for restarting the integration or the
small timestep start up costs for the integration method.

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}

@@

For the ODE integrations shown we use the variable-order,
variable-step integrator \vode\ \cite{vode} with its implementation of
the backward-differentiation formulas.
% The integration order
%corresponds to the order of the interpolating polynomials used to
%predict the solution in each integrator step. 
We note that for both
Strang splitting and SDC, \vode\ must start the integration at first
order and correspondingly small timesteps since no higher-order
information about the solution is available (our reaction
network right-hand sides provide only the first time derivative of the
solution). Based on testing with a 13-isotope
approximate alpha-chain network, \vode\ takes many small timesteps
(e.g.\ $\sim 200$) before it can stably increase to a high order
representation of the solution, but after it does so, it can often
complete the integration in only a handful of steps (e.g.\ $\sim
5$--$6$). Observing that during a \castro\ timestep, the first Strang
step is a direct continuation of the integration in the second Strang
step of the previous \castro\ timestep, we explored saving the
\vode\ state following the second Strang step and restarting
\vode\ during the first Strang step. This approach allowed us to start
the first Strang step with a high order representation of the solution
and complete it in very few \vode\ timesteps ($\lesssim 10$ steps),
speeding up the \castro\ timestep advance by 50\% for our detonation
test problem using the 13-isotope network. However, this came at the
memory cost of storing $7 N_{eq} + 90$ scalar variables on our grid,
where for our networks $N_{eq} = N_{spec} + 2$ (in addition to the
number of species $N_{spec}$ we also integrate temperature and
specific energy). For a large-scale science problem with a large
network, this could quickly become prohibitive, so we are exploring
ways to mitigate either the memory requirements for restarting the
integration or the small timestep start up costs for the integration
method.
